### 2020-03-03 doc/icfp20/conclusion.tex
fix typos

For now we have explicity converted between equivalent representations
and explictly lifted functions, to get a feel for the amount of work
required.  However in the world of dependent types there are known
research techniques for handling these cases, namely transporting
across equivalences as in Homotopy Type Theory~\citep{hottbook} and
ornaments~\citep{dagand-ornaments}. Cubical
Agda~\citep{vezzosi-cubical}, although still early in development,
should be an excellent tool for exploring the extent to which these
research ideas can be applied in a practical context.

Aside from these more fundamental ideas there is simply the work of
further extending the formalization of music theory to encompass
harmonic analysis, more advanced counterpoint, voice leading, and even
composition~\citep{schoenberg-fundamentals}. The results so far have
been encouraging, but much more needs to be done to find the right
abstractions to express the rules in the simplest, clearest and most
composeable way possible.
We expect the world of music theory to worthy of a long exploration.

>>>>

For now we have explicitly converted between equivalent representations
and explicitly lifted functions, to get a feel for the amount of work
required.  However in the world of dependent types there are known
research techniques for handling these cases, namely transporting
across equivalences as in Homotopy Type Theory~\citep{hottbook} and
ornaments~\citep{dagand-ornaments}. Cubical
Agda~\citep{vezzosi-cubical}, although still early in development,
should be an excellent tool for exploring the extent to which these
research ideas can be applied in a practical context.

Aside from these more fundamental ideas there is simply the work of
further extending the formalization of music theory to encompass
harmonic analysis, more advanced counterpoint, voice leading, and even
composition~\citep{schoenberg-fundamentals}. The results so far have
been encouraging, but much more needs to be done to find the right
abstractions to express the rules in the simplest, clearest and most
composable way possible.
We expect the world of music theory to be worthy of a long exploration.

### 2020-03-03 doc/icfp20/counterpoint.tex
fix typos

The middle bars of the music should maintain harmonical consonance
and independence of melodic lines.
As we saw before, consonant intervals include the unison,
3rd\footnote{We use ``3rd'' to mean both the major and minor variants
of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
and among these, the unison is clearly an obstacle to distinguishing
between the two lines of music.
Therefore, the middle bars must consist only of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors correponding to the occurrences of dissonant
intervals and unisons.

@@

The ending of the music should express relaxation.
There is a concensus that the unison and 8th are the most stable
intervals, hence the music must end with either of these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution
(Figure~\ref{fig:motioncadence} right).
This in turn suggests that a valid ending requires the middle bars
to be non-empty.
We encode these rules as the \texttt{checkEnding} function, which,
upon finding an invalid ending, reports one of the three possible
errors.

@@

Now we define \texttt{SecondSpecies}, a record type inhabited by
correct second species counterpoint.
As in \texttt{FirstSpecies}, we have three fields holding the musical
content, followed by five fields carrying the proofs of the required
properties disucssed above\footnote{
  The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}
  extracts the counterpoint note of a given interval, and
  \texttt{expandPitchIntervals2} turns a list of \texttt{PitchInterval2}s
  into a list of \texttt{PitchInterval}s.
}.

>>>>

The middle bars of the music should maintain consonance of intervals
and independence of melodic lines.
As we saw before, consonant intervals include the unison,
3rd\footnote{We use ``3rd'' to mean both the major and minor variants
of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
and among these, the unison is clearly an obstacle to distinguishing
between the two lines of music.
Therefore, the middle bars must consist only of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors corresponding to the occurrences of dissonant
intervals and unisons.

@@

The ending of the music should express relaxation.
There is a consensus that the unison and 8th are the most stable
intervals, hence the music must end with either of these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution
(Figure~\ref{fig:motioncadence} right).
This in turn suggests that a valid ending requires the middle bars
to be non-empty.
We encode these rules as the \texttt{checkEnding} function, which,
upon finding an invalid ending, reports one of the three possible
errors.

@@

Now we define \texttt{SecondSpecies}, a record type inhabited by
correct second species counterpoint.
As in \texttt{FirstSpecies}, we have three fields holding the musical
content, followed by five fields carrying the proofs of the required
properties discussed above\footnote{
  The auxiliary function \texttt{secondPitch} in \texttt{weakBeatsOk}
  extracts the counterpoint note of a given interval, and
  \texttt{expandPitchIntervals2} turns a list of \texttt{PitchInterval2}s
  into a list of \texttt{PitchInterval}s.
}.

### 2020-03-03 doc/icfp20/harmony.tex
fix typos

\citet{magalhaes-harmtrace} aim to creating a grammar for
harmonic analysis and a corresponding parser that takes as input a
series of chords and tries to produce a harmonic analysis that fits
the progression. The grammar makes crucial use of dependent types and
is thus awkard to represent in Haskell. We have implemented those
parts in Agda (\texttt{Fmmh.agda}) and not surprisingly they are simple and
elegant. However \citet{magalhaes-harmtrace} also makes use of
sophisticated Haskell libraries for parser combinators and generics,
which are unavailable in Agda and would either have to be natively
implemented or called through Agda's Haskell FFI, both of which involve
considerable work, so we have not yet pursued that route.

@@

Their techinique is to first require the melody notes to be members of
the harmonizing chords, then use their previous work to try to
create a harmonic analysis of every possible sequence of chords that
arises, chosing the one that most closely fits their
grammar. Although the result is not bad, it clearly leaves room for
improvement. Noted in the paper is that no attention is given to
voicing (how each of the four parts flows horizontally and how the
lines interact---in other words the counterpoint). Not noted is that
the melody note is always doubled, whereas typically in four-part
harmony it is prefered to double the root note. Also mimimizing the
number of parse errors as a means to choose the best harmonization
does not necessarily have a musical meaning.

Not having an impelmentation of the grammar and parser we explore a
different approach, based closer on Piston's.

>>>>

\citet{magalhaes-harmtrace} aim to creating a grammar for
harmonic analysis and a corresponding parser that takes as input a
series of chords and tries to produce a harmonic analysis that fits
the progression. The grammar makes crucial use of dependent types and
is thus awkward to represent in Haskell. We have implemented those
parts in Agda (\texttt{Fmmh.agda}) and not surprisingly they are simple and
elegant. However \citet{magalhaes-harmtrace} also makes use of
sophisticated Haskell libraries for parser combinators and generics,
which are unavailable in Agda and would either have to be natively
implemented or called through Agda's Haskell FFI, both of which involve
considerable work, so we have not yet pursued that route.

@@

Their technique is to first require the melody notes to be members of
the harmonizing chords, then use their previous work to try to
create a harmonic analysis of every possible sequence of chords that
arises, choosing the one that most closely fits their
grammar. Although the result is not bad, it clearly leaves room for
improvement. Noted in the paper is that no attention is given to
voicing (how each of the four parts flows horizontally and how the
lines interact---in other words the counterpoint). Not noted is that
the melody note is always doubled, whereas typically in four-part
harmony it is preferred to double the root note. Also minimizing the
number of parse errors as a means to choose the best harmonization
does not necessarily have a musical meaning.

Not having an implementation of the grammar and parser we explore a
different approach, based closer on Piston's.

### 2020-03-03 doc/icfp20/intro.tex
fix typos

Starting in the 20th century, computers have become ubiquitous in
music in every area, including sound sythesis, composition and
production~\citep{roads-tutorial}. In terms of music theory, there has
been a line of recent work on using functional programming for
harmonic
analysis~\citep{magalhaes-harmtrace,dehaas-harmtrace-a,dehaas-harmtrace-b},
harmonization of a melody and generation of melodies based on a
harmonization~\citep{koops-fharm,magalhaes-fcomp} and
counterpoint~\citep{szamozvancev-welltyped}. There is also an
established Haskell library Euterpea~\citep{hudak-haskell} for general music
and sound exploration.

@@

In this paper we explore what can be done in the context of music by
using a programming language that offers full dependent types. We use
Agda~\citep{norell-phd} since it is fairly mature and
aims to be both a functional programming language and a proof
assistant. It also features a Haskell FFI so we can take advantage of
existing Haskell libraries, in partiular for MIDI generation.

>>>>

Starting in the 20th century, computers have become ubiquitous in
music in every area, including sound synthesis, composition and
production~\citep{roads-tutorial}. In terms of music theory, there has
been a line of recent work on using functional programming for
harmonic
analysis~\citep{magalhaes-harmtrace,dehaas-harmtrace-a,dehaas-harmtrace-b},
harmonization of a melody and generation of melodies based on a
harmonization~\citep{koops-fharm,magalhaes-fcomp} and
counterpoint~\citep{szamozvancev-welltyped}. There is also an
established Haskell library Euterpea~\citep{hudak-haskell} for general music
and sound exploration.

@@

In this paper we explore what can be done in the context of music by
using a programming language that offers full dependent types. We use
Agda~\citep{norell-phd} since it is fairly mature and
aims to be both a functional programming language and a proof
assistant. It also features a Haskell FFI so we can take advantage of
existing Haskell libraries, in particular for MIDI generation.
